Approximate bayesian computation in large scale structure modeling : constraining the galaxy-halo connection

### 2016-01-19 halo/text/abc.tex
more text

\newcommand{\todo}[1]{{\bf \textcolor{red}{ #1}}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\lang}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\vep}{\bm{\epsilon}}
\newcommand{\ep}{\epsilon}

\begin{document}

\title{Likelihood free inference of the halo model parameters with group and clustering 
       statistics of the simulations of galaxy mock catalogs}

\begin{abstract}
The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies
populate dark matter halos that has had much success in reproducing large scale galaxy
clustering and understanding galaxy evolution within the context of the halo model.
Constraints on the parameters that describe the HOD model have so far been inferred using 
the standard practice of Bayesian inference with likelihood estimation that assumes a
functional Gaussian form of the likelihood.\todo{(Do we want to include analytic models for clustering statistics here?)} Likelihood free inference methods such as the
Approximate Bayesian Computation (ABC) makes Bayesian inference possible without ever 
having to calculate the likelihood.

In this work, we present ABC in the context of constraining the parameters of the HOD model.
For a mock galaxy catalog created with HOD parameters selected from literature
\todo{(Should we specify Zheng07?)}, we use ABC on measurements of a set of observables
generated through {\it Halotools}, a publicly available \todo{insert description about
forward modeling of galaxy positions}, to infer precise constraints that reproduce the 
known HOD parameters.

In this work, we use ABC in order to infer the HOD parameters by assuming a prior over the parameters, and a distance function between 
statistical summaries of the data and those of the simulations.
Using a mock galaxy catalog created with known HOD parameters, we obtain precise constraints on the HOD parameters that 
reproduces the summary statistics of the mock galaxy catalog accurately.

Furthermore, since ABC does not require computing the likelihood, we include observables
beyond galaxy clustering such as the galaxy number density and the galaxy group 
multiplicity function, for which accurately constructing a likelihood is intractable. We demonstrate that adding these measurements to our set of observables leads to stronger constraints on the HOD parameters, 
 specifically the ones that govern the satellite galaxy population. Moreover, the constraints we infer that reproduce the known HOD parameters illustrate the success of ABC in the context of constraining HOD parameters.

Since ABC does not require computing the likelihood, we include observable beyond galaxy clustering such as 
the galaxy number density and the galaxy group multiplicity function. 
We demonstrate that adding these measurements leads to stronger constraints on the HOD parameters, specifically the parameters 
that control the population of the satellite galaxies.

@@

\todo{Hogg writes here ...}

Inference of the cosmological parameters and understanding the formation of galaxies within the context of large scale 
structure, requires understanding of the Galaxy-Halo connection. Halos are biased tracers of the large scale structure 
and galaxies reside in dark matter Halos. Halo occupation distribution---HOD---is a probabilistic framework that describes how galaxies populate dark matter halos. The central tenet of halo occupation model is that the probability that a certain number of galaxies populate a dark matter halo is determined by the mass of the host halo. This statistical prescription for connection of galaxies and halos has been successful in reproducing the galaxy two-point clustering as well as the galaxy-matter cross correlation---galaxy-galaxy 
lensing---within the context of the halo model, and has been used in understanding the co-evolution of galaxies and halos and putting constraints on the relation between the stellar mass and halo mass of galaxies.

In the context of large scale clustering of galaxies, inference of the cosmological and halo model parameters are performed assuming an analytic model for the two-point correlation function of the galaxies, and a Gaussian likelihood. The accuracy of the analytic model of the galaxy clustering statistics hinges upon the accuracy of the fitting functions such as halo bias, halo mass function that are accurate to ten percent level. There are other simplifying assumptions encoded in these analytic models such as the scale independence of the halo bias.

Furthermore, the likelihood of observing the data---observed positions of the galaxies in the configuration space---given the cosmological and halo model parameters is approximated by a Gaussian function whose argument is the squared difference of the two-point correlation function measured from the observations and the analytic model for the galaxy clustering, weighted by a covariance matrix. The weighted difference of the number density of the observed galaxies and the analytic number density is also added to this. In this work, we investigate the how accurately we can infer the model parameters by relaxing the underlying assumptions of the parameter inference methods.

@@

ABC is a likelihood free method for inference of a set of parameters that are required for the simulation of a 
physical system, through a forward model, a distance function, and a prior probability distribution over the parameters. 
Let us denote the the data by $\mathcal{D}$. We assume that there exist a forward model $\mathcal{M}$ that 
simulates the data $\bar{\mathcal{D}}$ for a given set of parameters $\bm{\theta}$
\beq
\bar{\mathcal{D}}_{\theta} = \mathcal{M}(\bm{\theta})
\eeq

Furthermore, ABC requires adopting a distance function between the data $\mathcal{D}$ and the simulation $\bar{\mathcal{D}}_{\theta}$.
The distance function, in general, can be a vector $\bm{d}(\mathcal{D},\bar{\mathcal{D}}_{\theta})$
\begin{eqnarray}
\bm{d}(\mathcal{D},\bar{\mathcal{D}}_{\theta}) &=& [d_{1,\theta}, ... , d_{M,\theta}], \\
d_{i,\theta} &=& d_{i}(\mathcal{S}_{i}(\mathcal{D}), \mathcal{S}_{i}(\bar{\mathcal{D}}_{\theta}))  ~~\mbox{for $i=\{1,...,M\}$},
\end{eqnarray}
where $d_{i}(\dot,\dot)$ is a scalar function that defines a distance 
between $i$-$th$ summary statistics of the data $\mathcal{S}_{i}(\mathcal{D})$ and the same summary statistics of 
the simulation $\mathcal{S}_{i}(\bar{\mathcal{D}}_{\theta})$, and $M$ is the total number of statistical summaries 
used in the analysis.

We draw a large number of proposals from the prior distribution over the parameter space, 
and reject the samples for which the distance between the data and the simulation corresponding to that 
sample is more than a certain threshold $\bm{\epsilon}$. This requirement is satisfied \emph{if and only if}
every component of the distance vector is less than its corresponding threshold. Note that, each summary statistics 
has a separate threshold. This procedure is done iteratively untill, we are left with a certain number
of samples that satisfy the distance criteria. The final ensemble approximates the posterior probability distribution over the parameters of the generative model of the data. This sampling method is called \emph{rejected sampling}.

Rejected sampling is expensive. Instead we use Population Monte Carlo, adaptive importance sampling, ....

\subsection{Halo Model}

Halo Occupation distribution is a powerful tool for description of galaxy/halo relation. \todo{Here we explain HOD modeling of galaxy positions, etc.}

...

\subsection{Summary statistics}
\begin{itemize}

\item \todo{Talk about the observables: 2pcf, group multiplicity function, and number density ...}

\end{itemize}

\section{Forward modeling of the galaxy population with HALOTOOLS}\label{sec:halo}

\begin{itemize}
\item \todo{Halotools allows us to forward model galaxy positions from HOD parameters. From the galaxy positions, we can calculate observable summary statistics such as the 2PCF, Group Multiplicity Function (GMF) and the number density $\bar{n}$.}

\end{itemize}
Talk about why it is better to simulate the mock and compute the summary statistics of the mock instead of 
using analytic formulas...

@@

Talk about why we do not want to write down likelihood function for gmf, 2pcf, ...

...

\subsection{Mock data}
Plots showing 2pcf, gmf, ... of the mock data.
How we compute the covariance matrix from mocks...
Talk about error bars. Talk about poisson errors in group populations ...

@@

\subsection{Comparison to MCMC}
\todo{Apples to apples comparison between ABC and MCMC.}
\begin{itemize}
\item ABC-PMC using nbar and xi vs MCMC
\cite{Ishida:2015aa}
\end{itemize}

>>>>

\newcommand{\todo}[1]{{\bf \textcolor{red}{ #1}}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\lang}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\vep}{\bm{\epsilon}}
\newcommand{\ep}{\epsilon}
\newcommand{\pars}{\vec{\theta}}
\newcommand{\dev}{\mathrm{d}}

\begin{document}
% We need a better definitely shorter title...
\title{Approximate bayesian computation in large scale structure modeling : constraining the galaxy-halo connection}
\begin{abstract}
\todo{This is not a paper about hod, let's not start with describing hod. I suggest that we talk about parameter estimation in large scale structure in general and then move on to hod. Constraining hod is a just the first step in testing our methodology.} The Halo Occupation Distribution (HOD) model is a powerful prescription of how galaxies
populate dark matter halos that has had much success in reproducing large scale galaxy
clustering and understanding galaxy evolution within the context of the halo model.
Constraints on the parameters that describe the HOD model have so far been inferred using 
the standard practice of Bayesian inference with likelihood estimation that assumes a
functional Gaussian form of the likelihood.\todo{(Do we want to include analytic models for clustering statistics here? Yes, also we should say we use summary statistics that are not independent: clustering + group. Also there is no analytic model for the abundance of galaxy group richnesses)}  % I think we should keep it simple. There's no need to go into detail about everything in the abstract. This is already a pretty sizable abstract. 
Likelihood free inference methods such as the
Approximate Bayesian Computation (ABC) makes Bayesian inference possible without ever 
having to calculate the likelihood.

In this work, we present ABC in the context of constraining the parameters of the HOD model
using {\it Halotools}, a publicly available ... First, we set a mock galaxy catalog created
with HOD parameters from \cite{Zheng:2007aa} \todo{why do we site here?} as our data catalog. Then we use ABC on a set 
% To give it a false sense of legitimacy... 
of observables measured from forward modeled mock catalogs generated with {\it Halotools} 
in order to accurately infer constraints on the HOD parameters that 
reproduce the original parameters. \todo{Remark about the comparison between our ABC 
method constraints and the standard MCMC constraints.} Furthermore, since ABC does 
not require computing the likelihood, we include observables beyond galaxy clustering 
such as the galaxy number density and the galaxy group multiplicity function, for which
accurately constructing a likelihood is intractable. 
We demonstrate that adding these measurements to our set of observables leads to stronger
constraints on the HOD parameters, specifically the ones that govern the satellite galaxy
population. Moreover, the constraints we infer illustrate the feasibility of ABC in 
cosmological contexts.

@@

Cosmology was revolutionized in the 1990s with the introduction of likelihoods---%
pro\-ba\-bil\-iites for the data given the theoretical model and parameter settings---%
for combining data from different surveys and performing principled inferences of
the cosmological parameters (CITE CITE).
Nowhere has this been more true than in cosmic microwave background studies,
where it is nearly possible to analytically evaluate a likelihood function that
involves no (or minimal) approximations (CITE Wandelt and others).

Fundamentally, the tractability of likelihood functions in cosmology flows from
the fact that the initial conditions are exceedingly close to Gaussian in form,
and many sources of measurement noise are also Gaussian.
Likelihood functions are easier to write down and evaluate when things are closer to Gaussian,
so they get easier as the length scales get larger and the structure formation gets
younger.
Hence the suitability of CMB data for likelihood analyses.

In large-scale structure studies, which have galaxies, quasars, or quasar absorption
systems as tracers, the likelihood cannot be Gaussian:
Even if the initial conditions are perfectly Gaussian, the growth of structure
creates non-linearities which are non-Gaussian, and the galaxies form
within the density field in some complex manner that is unknown (or modeled only effectively).
Even if the galaxies were a Poisson sampling of the density field (which they are demonstrably
not; CITE PEOPLE), it would be tremendously hard to write down even an approximate
likelihood function.

The standard approach is to make two strong assumptions, both demonstrably wrong:
The first is that the (density and) two-point function measurements constitute
sufficient statistics for the cosmological parameters.
That is, the assumption is that all of the information about the cosmological model is
encoded in the mean and variance statistics of the tracer population.
This first assumption is equivalent to assuming that the density field is a Gaussian process,
and is therefore closer to true on larger scales.

The second strong and wrong assumption is that the likelihood functon for the data can
be approximated by a pseudo-likelihood function that is a Gaussian probability density
in the space of the two-point correlation function estimate.
That is, the assumption is that a pseudo-likelihood of the data given by a squared distance
(scaled by an inverse covariance matrix) between an estimate of the correlation function
(the data) and a theoretical prediction of that same function will return sensible posterior
inferences.
This assumption cannot be correct (in detail) at any scale, since a correlation function,
being related to the variance of a continuous field,
must satisfy non-trivial positive-definiteness
requirements (it must take the form of a valid kernel function; CITE SOMETHING).
These requirements truncate function space such that the likelihood in that functon space
could never be Gaussian.
This second assumption becomes truer as the correlation function gets better measured, so
it is most true on intermediate scales (where neither shot noise nor cosmic variance blow up).
We expect it to be very strongly violated at the baryon acoustic scale in particular!

Neither of these two assumptions are required for cosmological inferences, however:
We have very sophisticated simulations and emulations of structure and galaxy formation.
Indeed, the theory of large-scale structure \emph{is} the output of these simulations;
we can simulate the data in great detail.
Therefore, we can simulate not just the one- and two-point statistics of the galaxies,
but also any higher order statistics that might be relevant.
Furthermore, we can see the distributions of these statistics, and where the data lie within
those distributions.
That is---in principle---we don't need to make these strong and wrong assumptions;
we can create---in principle---a likelihood function directly from (an enormous number of)
simulation outputs.

Of course, any naive approach to sufficiently simulating the data would be ruinously
expensive.
Fortunately, there are principled, (relatively) efficient methods for minimizing computation
and delivering correct posterior inferences,
using only a data simulator and some choices about statistics.
Here we use \todo{Introduce the names ABC and LFI with references}.

As we will see, the LFI approach---despite cleverness---will be very computationally expensive.
For context, it is worth remembering that even the traditional approach to large-scale structure,
outlined above, requires enormous numbers of structure- and galaxy-formation simulations as well,
to construct the covariance matrices that go into the pseudo-likelihood.
So although our costs will be high, there are no inexpensive methods to compare to.

\todo{MJ: What Hogg wrote above doesn't mesh with (that is, repeats differently) what you have written below, so can you go through and fix that?  Lots of what is below should remain, but some should get woven into what's above.}

% style note: Here and in what follows: --- separates parenthetical phrases, -- connects words that appear together in a relationship (like galaxy--galaxy) and (galaxy--halo), and - connects pairs of words into an adjective (like two-point).

Inference of the cosmological parameters and understanding the formation of galaxies within the context of large scale 
structure, requires understanding of the galaxy--halo connection. Halos are biased tracers of the large scale structure 
and galaxies reside in dark matter Halos. Halo occupation distribution---HOD---is a probabilistic framework that describes how galaxies populate dark matter halos. The central tenet of halo occupation model is that the probability that a certain number of galaxies populate a dark matter halo is determined by the mass of the host halo. This statistical prescription for connection of galaxies and halos has been successful in reproducing the galaxy two-point clustering as well as the galaxy--matter cross correlation---galaxy--galaxy 
lensing---within the context of the halo model, and has been used in understanding the co-evolution of galaxies and halos and putting constraints on the relation between the stellar mass and halo mass of galaxies.

In the context of large scale clustering of galaxies, inference of the cosmological and halo model parameters are performed by treating some estimate of the galaxy--galaxy two-point correlation function as the data,
assuming an analytic model for the theoretical two-point correlation function prediction,
and a Gaussian likelihood connecting the two.
The accuracy of the analytic model of the galaxy clustering statistics hinges upon the accuracy of the fitting functions such as halo bias, halo mass function that are accurate to ten percent level. There are other simplifying assumptions encoded in these analytic models such as the scale independence of the halo bias.

Furthermore, the likelihood of observing the data---observed positions of the galaxies in the configuration space---given the cosmological and halo model parameters is approximated by a Gaussian function whose argument is the squared difference of the two-point correlation function measured from the observations and the analytic model for the galaxy clustering, weighted by a covariance matrix.
\todo{Why MUST this be wrong?}
The weighted difference of the number density of the observed galaxies and the analytic number density is also added to this. In this work, we investigate the how accurately we can infer the model parameters by relaxing the underlying assumptions of the parameter inference methods.

@@

ABC is a generative simulation-based inference based on rejection sampling. That is, the parameters of the forward model of the data are drawn from the prior probability, and are rejected if the simulated data is not sufficiently close to the observations according to some distance metric $\rho$. The method requires a prior probability over the parameters of interest $\pi(\theta)$, a generative forward model of the data given a set of parameters $f(\theta)$, and a distance metric $\rho(.,.)$ which measures the distance between the simulation $X=f(\theta)$ and the observed data $D$.

In a large scale structure study however, the observed data---which encompasses the angular positions of galaxies on the sky and their redshifts---is a high dimensional object. Therefore the distance metric measures the closeness of statistical summaries of the data and those of the simulation. When the observed data-set under study is the positions of all galaxies observed in the survey, possible statistical summaries of the data are the number density of galaxies, two-point correlation function, three-point correlation function, galaxy group statistics (counts in cells), and so on.

Here, we provide a brief overview of rejection sampling algorithm and how it is used in a simulation-based inference. We want to draw samples from a probability distribution $p(\alpha)$. Now, let us assume that: (i) there exists an auxiliary density $q(\alpha)$ from which we can directly sample from, and (ii) there exist $M>1$ which is an appropriate bound on $p(\alpha)/q(\alpha)$. Given these assumptions, rejection sampling works in the following way. We draw a sample $\alpha$ from the auxiliary density $q(\alpha)$. Afterwards, we draw a random number $u$ from a uniform distribution between zero and one. Then we accept the draw $\alpha$ if the $u$ is less than the ratio $p(\alpha)/Mq(\alpha)$. This procedure is repeated until sufficient number of draws are accepted.

In the context of a simulation-based inference, we aim to sample from the joint probability of a simulation $X$, and parameters $\pars$ given the observed data $D$

\beq
p(\pars,X | D)
\eeq

which according to the Bayes rule can be written as

\begin{eqnarray}
p(\pars,X | D)  &=& \frac{p(D|X)p(X|\pars)\pi(\pars)}{Z},  \\
Z &=& \int \dev \pars \; \dev X \; p(D|X)p(X|\pars)\pi(\pars),
\end{eqnarray}
where $Z$ is the evidence. There exists a constant $M>1$ 
\beq 
M = \frac{\mathrm{max}_{X} \; p(D|X)}{Z},
\eeq 
and auxiliary density $q(\pars ,X)$
\beq 
q(\pars ,X)=p(X|\pars )\pi(\pars ),
\eeq 
such that:
\beq 
p(\pars ,X|D)<Mq(\pars ,X).
\eeq 
Now in a simulation-based inference, we first draw a pair $\{\pars , X \}$ from the density $q(\pars , X)$. This requires drawing a sample $\pars$ from the prior $\pi(\pars)$, and generating a simulation $X=f(\pars)$ via the forward model. Then, we accept the pair $\{\pars , X \}$, if the following condition is met

\beq 
u \sim \mathrm{Uniform}[0,1] < \frac{p(\pars,X|D)}{Mq(\pars,X)}=\frac{p(D|X)}{\mathrm{max}_{X}p(D|X)}.
\eeq

ABC postulates that the probability of observing data $D$ given simulation $X$, $p(D|X)$ is controlled by a threshold $\epsilon$, and that this probability is one if the distance between the data and the simulation, $\rho(D,X)$ is less than $\epsilon$. That is

\beq 
p(D|X) = p_{\epsilon}(D|X) \equiv \begin{cases}
                                    1 \ \ \  \mathrm{if} \ \rho(D,X)<\epsilon \; , \\
                                    0 \ \ \ \mathrm{otherwise} \; .
                                    \end{cases}
\eeq

Thus the set of parameters $\pars$ is accepted if $\rho(D,f(\pars))<\epsilon$. In principle, the distance function measures the distance 
between the statistical summaries of the data and those of the simulation. 
The distance, in general, can be a vector, with each component representing the distance between one summary statistics of the data (e.g. variance) and the same summary statistics of the simulation. That is
\beq 
\rho(D,X) = \Big[\rho_{1}(S_{1}(D) , S_{1}(X)), ... , \rho_{M}(S_{M}(D) , S_{M}(X)) \Big],
\eeq 
where $M$ is the number of statistical summaries used in the inference, $S_{i}$ is the i-th summary statistics, and $\rho_{i}$ is a scalar function that measures the distance between the i-th summary statistics of the data and that of the simulation. In this case, the threshold $\epsilon$ is an $M$ dimensional vector.

We draw a large number of proposals from the prior distribution over the parameter space, 
and reject the samples for which the distance between the data and the simulation corresponding to that 
sample is more than a certain threshold $\bm{\epsilon}$. This requirement is satisfied \emph{if and only if}
every component of the distance vector is less than its corresponding threshold. Note that, each summary statistics 
has a separate threshold. This procedure is done iteratively until, we are left with a certain number
of samples that satisfy the distance criteria. The final ensemble approximates the posterior probability distribution over the parameters of the generative model of the data.

For a sufficiently small threshold, this algorithm samples from the posterior probability, $p(\pars | D)$, and for an arbitrarily large threshold, the algorithm samples from the prior $\pi(\pars)$. This rejection sampling technique requires repeated sampling from the prior, and a very small threshold to succeed.

In this investigation instead, we use Population Monte Carlo Algorithm (PMC) as our sampling technique. PMC is an iterative method that performs rejected sampling from a sequence of distributions $\{p_{1}(\pars),...,p_{T}(\pars)\}$.These distributions are paired with a sequence of decreasing distance thresholds $\{\epsilon_{1} , ... , \epsilon_{T}\}$.

At each iteration, $N$ proposals are drawn from a distribution over parameters, and a weight is assigned to each one. In the first step, the threshold is set to an arbitrarily large value. Therefore the algorithm makes $N$ draws from the prior and assigns equal weights to all the proposals.

In the consequent iterations, the distance threshold is set to a smaller value. A proposal is weighted-sampled from the previous sample. It is perturbed by a kernel which is set to the covariance of the parameters in the previous sample. A simulation is then generated by feeding the perturbed parameter into the forward model. The perturbed parameter is then accepted if the distance between the data and the generated simulation is less than the value of the threshold in that iteration. This procedure is repeated until $N$ particles are sampled. Then the weights are updated according to the kernel, the prior, and the previous set of weights.

\subsection{Forward model of the large scale galaxy populations}\label{sec:halo}
\subsubsection{halo model}
Halo Occupation distribution is a powerful tool for description of galaxy/halo relation. \todo{Here we explain HOD modeling of galaxy positions, etc.}
\subsubsection{Generating mock galaxy populations with halotools}

\todo{Halotools allows us to forward model galaxy positions from HOD parameters. From the galaxy positions, we can calculate observable summary statistics such as the 2PCF, Group Multiplicity Function (GMF) and the number density $\bar{n}$.}

Talk about why it is better to simulate the mock and compute the summary statistics of the mock instead of 
using analytic formulas...

@@

Talk about why we do not want to write down likelihood function for gmf, 2pcf, ...
\subsubsection{summary statistics of galaxy populations}

\subsubsection{distance metrics}

\todo{Talk about the observable: 2pcf, group multiplicity function, and number density ...}

\section{Mock data}\label{sec:data}

Plots showing 2pcf, gmf, ... of the mock data.
How we compute the covariance matrix from mocks...
Talk about error bars. Talk about poisson errors in group populations ...

@@

\subsection{Comparison to MCMC}
\todo{Apples to apples comparison between ABC and MCMC.}
ABC-PMC using nbar and xi vs MCMC
